**Density Dependence**

### 2023-03-10 Rmd/report whb.Rmd
read thru

---
output:
  word_document:
    reference_docx: ../report_template_v1.1.dotx
mathjax: TRUE
bibliography: refs.bib
---

@@

All conditioning of the operating models and estimation of potential density dependent processes have been based on the ICES 2022 SAM assessment of `r mystkname`.

@@

The operating models for this study are derived from the 2022 stock assessment. SAM is a statistical catch-at-age model that estimates error in the catch-at-age. In the case of `r mystkname`, the assessment gives a higher weight to the recruitment index than to the catch data for ages 0 and 1. This means that the estimates of recruitment in the assessment can be very different from the actual number of fish that are recruited to the fishery. To ensure the MSE projections and the historical estimates are comparable, a Virtual Population Analysis (VPA) was conducted on the whole catch matrix and the survivors in the most recent year and oldest age. In the next step, we applied a natural mortality-at-age vector derived from the Lorenzen approach which would mimic the VPA assessment with fixed M-at-age. The figure below compares the SAM assessment with the VPA and the VPAM assessment. A main difference is in the recruitment estimates. Using age-varying M tends to flatten the recruitment pattern. The vpaM model was used as the base case to be compared against the density-dependent simulations.

@@

Estimating density-dependence from observations was only possible for mass-at-age and maturity-at-age. Density dependence in natural mortality can only be explored from theoretical considerations (because current assessment does not include age or time varying natural mortality).

In the next step, we applied a natural mortality-at-age vector based on [@lorenzen2000allometry] which would mimic the VPA assessment with fixed M-at-age. The figure below compares the SAM assessment with the VPA and the VPAM assessment. A main difference is in the recruitment estimates. Using age-varying M tends to flatten the recruitment pattern. The vpaM model was used as the base case to be compared against the density-dependent simulations.

Next, the operating models were constructed. This was done in a number of steps.

@@

2.  Next, a VPA model model was estimated. This was done to ensure the MSE projections and the historical estimates are comparable. The Virtual Population Analysis (VPA) was conducted on the whole catch matrix and the survivors in the most recent year and oldest age.

3.  The third step was to convert the VPA model with fixed natural mortality at age and by year into an age-varying natural mortality vector. We applied a natural mortality-at-age vector derived from the Lorenzen approach (@lorenzen2000allometry) which would mimic the VPA assessment with fixed M-at-age. This assessment was then treated as the base case against which the density dependent operating models were compared.

@@

-   VPA Base Case, with varying M-at-age
-   Density dependence in mass-at-age,
-   Density dependence in mass, maturity-at-age,
-   Density dependence in mass, maturity and M-at-age,

The MSY reference points $B_{MSY}$ and $F_{MSY}$ were calculated based on equilibrium assumptions, either from combining yield per recruit and spawner per recruit relationship [@sissenwine1987alternative] or by conducting long-term projections under constant fishing moralities.

@@

The OMs were also projected for the estimates of different estimates $F_{MSY}$, with stochasticty in the recruitment deviates. This allows the impact of misspecification of density dependence to be evaluated, i.e. if $F_{MSY}$ is underestimated then yield will be less than MSY, while if $F_{MSY}$ is overestimated yield will be lost due to overfishing. Stochastic projections of the four Operating Models for the four estimates of $F_{MSY}$ are shown in **Figures 16-19**

**Figure 20** summarizes the yields and SSBs obtained for each Operating Model under the correct and misspecified form of density dependence and the associated reference points, and **Figure 21** shows the Kobe plots of the final 2050 distributions of biomass and yield.

Estimating density-dependence from observations was only possible for mass-at-age and maturity age. Therefore, density dependence in natural mortality can only be explored from theoretical considerations, because current assessment does not include age or time varying natural mortality. M is likely to vary by age and time in relation to processes such as density dependence. While the recent expansion in distribution area and change in weight-at-age could be associated with changes in natural mortality. Therefore in the OMs we modeled M as a function of mass-at-age.

@@

[\@ @This value was calculated based on estimates of total mortality derived from tagging data combined with catch data.]

>>>>

---
output:
  html_document:
    reference_docx: ../report_template_v1.1.dotx
mathjax: TRUE
bibliography: refs.bib
---

@@

All conditioning of the Operating Models and estimation of potential density dependent processes have been based on the ICES 2022 SAM assessment of `r mystkname`.

@@

The Operating Models for this study are derived from the 2022 stock assessment. SAM is a statistical catch-at-age model that estimates error in the catch-at-age. In the case of `r mystkname`, the assessment gives a higher weight to the recruitment index than to the catch data for ages 0 and 1. This means that the estimates of recruitment in the assessment can be very different from the actual number of fish that are recruited to the fishery. To ensure the MSE projections and the historical estimates are comparable, a Virtual Population Analysis (VPA) was conducted on the whole catch matrix and the survivors in the most recent year and oldest age. In the next step, we applied a natural mortality-at-age vector derived from the Lorenzen approach which would mimic the VPA assessment with fixed M-at-age. The figure below compares the SAM assessment with the VPA and the VPAM assessment. A main difference is in the recruitment estimates. Using age-varying M tends to flatten the recruitment pattern. The vpaM model was used as the base case to be compared against the density-dependent simulations.

@@

Estimating density-dependence from observations was only possible for mass-at-age but not maturity-at-age as the latter were assumed by ICES not to have varied over time. However, assuming that maturity is a function of mass-ata-ge allowed density deoendence to be modeeled. Natural mortality was assumed to be time and age invariant by ICES (i.e. 0.2), therefore density dependence in natural mortality can only be explored from theoretical considerations.

In the next step, we applied a natural mortality-at-age vector based on [@lorenzen2000allometry] which mimics the VPA assessment with fixed M-at-age. The figure below compares the SAM assessment with a Virtual Population Analysis (VPA) and a VPA with age varying M (VPAM). Trends and absolute levels swere similar across all runs. The vpaM model was used as the base case to be compared against the density-dependent simulations.

Next, the Operating Models were constructed. This was done in a number of steps.

@@

2.  Next, VPA was conducted. This was done to ensure the MSE projections, which assumed variable M-at-age, and the historical estimates are comparable. The VPA was conducted on the whole catch matrix and the survivors in the most recent year and oldest age.

3.  The third step was to convert the VPA model with fixed natural mortality at age and by year into an age-varying natural mortality vector. We applied a natural mortality-at-age vector derived from the Lorenzen approach (@lorenzen2000allometry) which would mimic the VPA assessment with fixed M-at-age. This assessment was then treated as the base case against which the density dependent Operating Models were compared.

@@

-   Base Case, VPA with varying M-at-age
-   Density dependence in mass-at-age,
-   Density dependence in mass, maturity-at-age,
-   Density dependence in mass, maturity and M-at-age,

The MSY reference points $B_{MSY}$ and $F_{MSY}$ were calculated based on equilibrium assumptions, either from combining yield per recruit and spawner per recruit relationship [@sissenwine1987alternative] or by conducting long-term projections under constant fishing moralities. Both should agree and so can be used as a check.

@@

The OMs were also projected for the four estimates of  $F_{MSY}$, with stochasticty in the recruitment deviates. This allows the impact of misspecification of density dependence to be evaluated, i.e. if $F_{MSY}$ is underestimated then yield will be less than MSY, while if $F_{MSY}$ is overestimated yield will be lost due to overfishing. Stochastic projections of the four Operating Models for the four estimates of $F_{MSY}$ are shown in **Figures 16-19**

**Figure 20** summarizes the yields and SSBs obtained for each Operating Model under the correct and misspecified form of density dependence and the associated reference points, and **Figure 21** shows phase plots of the final 2050 distributions of biomass and yield. THese are for the Base Case, then density-dependence in mass, mass and maturity, and mass, maturity and natural mortality, projected for the four estimates of $F_{MSY}$. The correctly specified $F_{MSY}$ for each model is indicated by the red points and marginal densities.

The main diofferences are seen in $SSB/B_{MSY}$, if no density dependence is assumed then in the operating model the estimate of $F_{MSY}$ is the lowest, therefore projreting for the higher estimates of $F_{MSY}$ assuming density dependence results in an overfished stock. Yileld are the same for the two values of $F_{MSY}$ that assume no density-dependence on density dependence in mass-at-age only. In the other two cases yield is reduced but only by about 10%.

In the cases where density-dependence is assumed in the Operating Model, yields do not vary much, the main effect is seen in the estimates of $SSB/B_{MSY}$,

@@
